---
title: "Mathematical Modeling of Epidemics"
author: "Daniel Paliura"
date: "3/23/2021"
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE)
```

```{r sourcing, include=FALSE}
library('deSolve')

source('model.R')
source('output.R')
```

#### Contents

* [Introduction](#introduction)
* [Description of chosen SIRD model](#description-of-chosen-sird-model)
* [Research of influence of model parameters](#research-of-influence-of-model-parameters)
    + [Number of people](#number-of-people)
    + [Number of contacts per day](#number-of-contacts-per-day)
    + [Probability of infection](#probability-of-infection)
    + [Recovery and death probabilities](#recovery-and-death-probabilities)
* [Vaccination control research](#vaccination-control-research)
* [My opinion](#my-opinion)

## Introduction

This document was created in addition to 'Control Theory' subject second
laboratory work 'Mathematical Modeling of Epidemics'. This document on GitHub
is [here](https://github.com/dPaliura/Control_theory_labs/blob/main/lab2/research.pdf)
It contains the research for a dependence of input parameters on the problem
solution. One of the well-known models for modeling epidemics is SIRD
(Susceptible, Infected, Recovered, Dead). All parameters will be set randomly in
recommended intervals that are respective to the COVID-19 pandemic. Also, the
SIRD model can be complicated with some special values.\

I chose the model with testing, so the system is described with additional value
J - Justified (I guess, it could mean something other). Justified - people who
tested with a positive result, in other words, people with proved infection.
And so I - Infected and latent people.\

This work refers to the control theory a little because the model contains
control parameters. It can be used to make research of pandemic change and also
to try different controls or strategies.

## Description of chosen SIRD model

Model is based on differential equations system, that contains each variable
change in time. SIRD with testing is described by the following system:
$$S(t)+I(t)+J(t)+R(t)+D(t) = N$$
$$\frac{dS}{dt} = -Sp(I) - Su(t)$$
$$\frac{dI}{dt} = Sp(I) - \alpha I - \beta I - Iv(t)$$
$$\frac{dJ}{dt} = Iv(t) - aJ - bJ$$
$$\frac{dR}{dt} = \alpha I + aJ + Su(t)$$
$$\frac{dD}{dt} = \beta I + bJ$$
with initial state $I(0)>0, S(0)=N-I(0)$, where\
$N$ - number of people;\
$S, I, J, R, D$ - number of Susceptible, Infected (latent), Infected (tested),
Recovered and Dead people at moment $t$;\
$p(t)dt = p(I(t))dt$ - probability to infect while time $dt$;\
$u(t) \in [0,1]$ - vaccinated part of people. Control variable,
by default $u(t)\equiv0$;\
$v(t) \in [0,1]$ - tested part of people at the moment $t$. Control variable;\
$\alpha$ - coefficient of recovery speed for latent infected;\
$\beta$ - mortality coefficient for latent infected;\
$a$ - coefficient of recovery speed for tested infected;\
$b$ - mortality coefficient for tested infected.\

Let $r$ is mean density of contacts for one person, $c$ - probability of
infection in case of contact with infected. So $\frac{I}{N}$ - probability to
meet infected. $\frac{rI}{N}dt$ - number of contacts with infected while time
$dt$. $1-c$ - probability not to infect while contact with infected.
$(1-c)^{\frac{rI}{N}dt}$ - probability not to infect while time $dt$. So
probability to infect while time $dt$ is:
$$1 - (1-c)^{\frac{rI}{N}dt} = 1 - \exp(\frac{r\log(1-c)I}{N}dt)
\approx 1-(1+\frac{r\log(1-c)I}{N}dt) = -\frac{r\log(1-c)I}{N}dt$$
So $p(t) = p(I(t)) = -\frac{rI\log(1-c)}{N}$.

To create a model of epidemic flow, there is a need to solve differential 
equations system with a numeric method. To do so I used R package ```deSolve```,
function  ```ode(y, times, func, parms, ...)```. Also, got values are decimals,
which is not correct to represent people, so I rounded $S, I$, floored $R, D$
and got $J = N-(S+I+R+D)$.

As model parameters will be taken:

* $N$ - number of people
* $I_0$ - number of infected people at moment $t=0$, $0<I_0<N$
* $r$ - mean number of contacts per day for a single person
* $c$ - probability of infection for single contact with infected
* $\alpha$ - probability to recover while one day for a single latent infected 
* $\beta$ - probability to die while one day for a single latent infected
* $a$ - probability to recover while one day for a single tested infected
* $b$ - probability to die while one day for a single tested infected

## Research of influence of model parameters

To do research I will variate each parameter with fixed others. I will set
uniform random values in recommended intervals as default:\
$N \in [10^3, 10^6], r \in [0.001, 50], c \in [0.4, 0.9]$\
$\alpha \in [0.05, 0.1], \beta \in [0.01, 0.1]$\
$a \in [\alpha, 0.1], b \in [0.01, \beta]$.\
Testing part function $v(t)$ will be as it was in real experience, something
like the cumulative function of normal distribution (first 20 days no tests,
then testing speed increasing fast and again decreasing).

```{r set default input, echo=TRUE}
set.seed(32)

input <- list(
    N = round(runif(1, 1e3, 1e6)),
    I0 = 1,
    r = 10^runif(1, log(0.001, 10), log(50,10)),
    c = runif(1, 0.4, 0.9),
    alpha = (runif(1, 0.05, 0.1) -> alpha),
    beta = (runif(1, 0.01, 0.1) -> beta),
    a = runif(1, alpha, 0.1),
    b = runif(1, 0.01, beta)
)

input <- lapply(input, round, 4)

print(input)

input$v = function(t) ifelse(test = t>20,
                             yes = 0.7*pnorm(t, 150, 100),
                             no = 0)
```

The built model is represented on the plot:

```{r default model build and plot}
model <- build.model(input)
display.model(model, 'Model with default parameters until end')
```

Looks controversial...

#### Number of people\

```{r N variate plots, fig.height=8}
par(mfrow=c(4,1))
par(mar = c(2,4,2,2))

input.tmp <- input

N <- 10^(3:6)
for (n in N){
    input.tmp$N <- n
    model <- build.model(input.tmp, 40)
    display.model(model, paste0('N = ', n), legend = FALSE)
}
```

All plots are almost similar, so $N$ affects weakly on epidemic flow. There is a
small shift: $I$ tops are at $t\approx10$ for $N=1e3$ and $t\approx17$ for
$N=1e6$, so the difference is scanty considering $N$ change.\
I conclude: **$N$ affects the speed of infection spread not significantly**.

#### Number of contacts per day\

```{r r variate plots}
input.tmp <- input

par(mfrow=c(3,2))
par(mar = c(2,4,2,2))

r <- c(.001, .01, .1, 1, 5, 50)
for (value in r){
    input.tmp$r <- value
    model <- build.model(input.tmp, 50)
    display.model(model, paste0('r = ', value), legend = FALSE)
}
```

Those plots show weak sides of the model. First of all: we see no infection
spread for $r \le 0.1$, it might be because the single infected person died
before  infecting someone other. This situation is very controversial. Also, a
lack of states is observed, because, for example, COVID-19 has a long incubation
term, while people can infect others but are not described as infected. And
also, I think intervals for parameters are given badly because the infection
spreads very fast in each example and also people begin to die from the first
day of infection. In my opinion, it isn't correct to say about probabilities to
recover or die ($\alpha, \beta, a, b$), when those probabilities are used in
determined the model. I think the word 'probability' is not compatible with
determined models. OK, also, we see that 
**increase of the $r$ parameter leads to faster infection spread**.
But limit distribution of Recovered and Dead almost the same, because
it's very few tested infected (testing starts too late). For $r=50$ there is
the situation that almost all people are infected on the second day.
**Contacts intensity highly affects on begin of infection spread**.

#### Probability of infection\

```{r c variate plots}
input.tmp <- input

par(mfrow=c(3,2))
par(mar = c(2,4,2,2))

c <- c(.4, .5, .6, .7, .8, .9)
for (value in c){
    input.tmp$c <- value
    model <- build.model(input.tmp, 150)
    display.model(model, paste0('c = ', value), legend = FALSE)
}
```
Very interesting. There is such an interesting situation: if the probability is
small, then there are non-zero rest if Susceptible people. I guess it's because
of testing: testing occupies latent Infected people before all Susceptible
become Infected, and it can be proved: on plots with $c \ge 0.7$ the testing
begins too late - when almost all Susceptibles become Infected. In other cases,
testing begins when Infection spread is not such rapid and does not occupy
almost all people. So according to these models, I conclude: 
**if testing exists and it is begun on time, it's available to stop infection**
**spread until all people infected in case infection probability is low**.
But there is an issue: it is not realistic, just demonstrative. And who knows
whether it's better to recover all people, or to stop infection. Probably we can
choose the strategy depending on this value. If the infection is low contagious,
then it should be isolated, because it's available, like for papilloma or AIDS.
But I mentioned sexually transmitted viruses. What about airborne droplets? I
think there are no ways to stop it because we have to make people not breathe.
For COVID-19 we have what we have, it has occupied the whole Earth.

#### Recovery and death probabilities\

I will compare equal sets of $\alpha$-s and $\beta$-s to get a more
understandable picture. I will also set $a=\alpha, b=\beta$.

```{r alpha and beta variate plots}
input.tmp <- input

par(mfrow=c(3,3))
par(mar = c(2,4,2,2))

alpha <- c(.01, .05, .1)
beta <- c(.01, .05, .1)
for (value1 in alpha){
    for (value2 in beta){
        input.tmp$alpha <- value1 -> input.tmp$a
        input.tmp$beta  <- value2 -> input.tmp$b
        model <- build.model(input.tmp, 100)
        
        main <- parse(text = paste0("paste(alpha, ",
                                    paste0("' = ", value1, ", ', "),
                                    "beta, ",
                                    paste0("' = ", value2, ", ', "),
                                    "alpha, '/', beta, ",
                                    paste0("' = ", value1/value2, "'"),
                                    ")"))
        
        display.model(model, 
                      main, 
                      legend = FALSE)
    }
}
```

All plots look similar, but there are tiny differences. Let's look at the
diagonal set of plots. We see equal distributions of Recovered and Dead but
different forms of lines. And on the last plot, there is very little rest of
Susceptible. Distributions of Recovered and Dead obey to $\frac{\alpha}{\beta}$
rate but, I noticed that height of Infected and hence Tested tops are different
and they are less when $\alpha+\beta$ greater and, also, lines becomes straight
earlier. I conclude: 
**if testing works nice and $a, b$ are better than $\alpha, \beta$, then there**
**is a sense to provide more testing (but we have to remember that hospitals**
**are not dimensionless)**.\
To prove this conclusion I shall demonstrate this. So let's see plots,
where $\alpha \in \{0.01, 0.05\}, \beta = 0.01$ and 
$a \in \{\alpha, 4\alpha\}, b \in \{\frac{\beta}{4}, \beta\}$,

```{r a and b variate plots}
input.tmp <- input

par(mfrow=c(2,2))
par(mar = c(2,4,2,2))

input.tmp$alpha <- 0.01 -> input.tmp$a
input.tmp$beta <- 0.01 -> input.tmp$b
model <- build.model(input.tmp, 200)
display.model(model, 
              expression(paste('a = ', alpha, ' = 0.01, b = ', beta,' = 0.01')), 
              legend = FALSE)

input.tmp$a <- 0.04
input.tmp$b <- 0.0025
model <- build.model(input.tmp, 200)
display.model(model, 
              expression(paste('a = 4',alpha,' = 0.04, b = ',beta,'/4 = 0.0025')), 
              legend = FALSE)

input.tmp$alpha <- 0.05 -> input.tmp$a
input.tmp$beta <- 0.01 -> input.tmp$b
model <- build.model(input.tmp, 200)
display.model(model, 
              expression(paste('a = ', alpha, ' = 0.05, b = ', beta,' = 0.01')), 
              legend = FALSE)

input.tmp$a <- 0.2
input.tmp$b <- 0.0025
model <- build.model(input.tmp, 200)
display.model(model, 
              expression(paste('a = 4',alpha,' = 0.2, b = ',beta,'/4 = 0.0025')), 
              legend = FALSE)
```

The first pair of plots is very demonstrative about how can testing affect death
volume. The second pair of plots example shows a simple idea:
**Testing is more sensible as sooner it begins**.

## Vaccination control research

Let try next vaccination controls: $u(t)=500/N$ and $u(t)=5000/N$ (500 and 5000
persons per day).


```{r}
par(mfrow=c(3,1))
par(mar = c(2,4,2,2))

input.tmp <- input


model1 <- build.model(input.tmp, 180)
display.model(model1,
              'u(t) = 0',
              legend=FALSE)

input.tmp$u <- function(t) 500/input$N
model2 <- build.model(input.tmp, 180)
display.model(model2,
              'u(t) = 500/N',
              legend=FALSE)

input.tmp$u <- function(t) 5000/input$N
model3 <- build.model(input.tmp, 180)
display.model(model3,
              'u(t) = 5000/N',
              legend=FALSE)


cat('u(t) = 0:\n')
M <- rbind(model1$states[nrow(model1$states),2:6],
           lapply(2:6, function(j) max(model1$states[,j])))
row.names(M) <- c('T', 'max')
print(M)

cat('u(t) = 500/N:\n')
M <- rbind(model2$states[nrow(model2$states),2:6],
           lapply(2:6, function(j) max(model2$states[,j])))
row.names(M) <- c('T', 'max')
print(M)

cat('u(t) = 5000/N:\n')
M <- rbind(model3$states[nrow(model3$states),2:6],
           lapply(2:6, function(j) max(model3$states[,j])))
row.names(M) <- c('T', 'max')
print(M)
```

There is a valuable difference only in the last case, but is it available to
make  5000 vaccinations per day in a half-million city? It's about 1% per day.
I guess it's about maximum because it can be a lack of staff to make such many
vaccinations per day. Only the last example shows notable changes in the plot.
Numbers differ for each case, but not such significant for first and second
cases as for second and third ones. I believe that human life is priceless and 
it is only the reason to vaccinate. Also, I think it makes sense to vaccinate on
epidemic's very beginning to provide a maximum of Recovered people and hence a
minimum of Infected to prevent many deaths.

## My opinion

I think this method is good for the fast inaccurate modeling, but not sufficient
to make some good realistic model, because this method gives a determined model.
I think such a process like an epidemic can't be determined, it has to be
stochastic. I guess it would be better to make something much complicated (even
complex to formulate) but stochastic. And use the Monte-Carlo method.
I tried to imagine what I would do if I would make my own model. Probably, it 
would be nice to keep numbers of Healthy (not infected), Infected (can be latent
and tested), Recovered and Immunized (aren't equal), and Dead too.
But the model has to be complicated with such parameters as:

* Incubation period mean duration
* Malady mean duration
* Quarantine properties (less contacts and probability of infection)
* Mean severity at the moment
* Severity deviation
* Social activity of people at weekdays and weekends
* Mean recovery times
* Probabilities to acquire immunity after recovery
* And others...

I imagine it like several capacities where can be people. Like in the Markov
chain, each person is in some group (Healthy/Infected/Dead...) but it's hard
to process each person with the Markov chain. So I propose to move random
amounts of people from each state (except Dead and Immunized) like in the Markov
chain but with consideration of mentioned parameters. For example, high severity
brings high death or higher mean recovery times decreases the probability to
infect again.
