---
title: "Mathematical Modeling of Epidemics"
author: "Daniel Paliura"
date: "3/23/2021"
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE)
```

```{r sourcing, include=FALSE}
library('deSolve')

source('model.R')
source('output.R')
```

## Introduction

This document was created as additional to 'Control Theory' subject second
laboratory work 'Mathematical Modeling of Epidemics'. This document contains
research for a dependence of input parameters on the problem solution. One of
well-known models for modeling epidemics is SIRD (Susceptible, Infected,
Recovered, Dead). All parameters will be set randomly in recommended intervals
that are respective to COVID-19 pandemic. Also, SIRD model can be complicated
with some special values.\

I chose model with testing, so system described with additional value
J - Justified (I guess, it could mean something other). Justified - people who
tested with positive result, in other words, people with proved infection.
And so I - Infected and latent people.\

This work refers to the control theory a little, because model contains control
parameters. It can be used to make research of pandemic change and also to try
different controls or strategies.

## Description of chosen SIRD model

Model is based on differential equations system, that contains each variable
change in time. SIRD with testing is described by following system:
$$S(t)+I(t)+J(t)+R(t)+D(t) = N$$
$$\frac{dS}{dt} = -Sp(I) - Su(t)$$
$$\frac{dI}{dt} = Sp(I) - \alpha I - \beta I - Iv(t)$$
$$\frac{dJ}{dt} = Iv(t) - aJ - bJ$$
$$\frac{dR}{dt} = \alpha I + aJ + Su(t)$$
$$\frac{dD}{dt} = \beta I + bJ$$
with initial state $I(0)>0, S(0)=N-I(0)$, where\
$N$ - number of people.\
$S, I, J, R, D$ - number of Susceptible, Infected (latent), Infected (tested),\
Recovered and Dead people at moment $t$.\
$p(t)dt = p(I(t))dt$ - probability to infect while time $dt$.\
$u(t) \in [0,1]$ - vaccinated part of people. Control variable,
by default $u(t)\equiv0$.\
$v(t) \in [0,1]$ - tested part of people at moment $t$. Control variable.\
$\alpha$ - coefficient of recovery speed for latent infected.\
$\beta$ - mortality coefficient for latent infected.\
$a$ - coefficient of recovery speed for tested infected.\
$b$ - mortality coefficient for tested infected.\

Let $r$ is mean density of contacts for one person, $c$ - probability of
infection in case of contact with infected. So $\frac{I}{N}$ - probability to
meet infected. $\frac{rI}{N}dt$ - number of contacts with infected while time
$dt$. $1-c$ - probability not to infect while contact with infected.
$(1-c)^{\frac{rI}{N}dt}$ - probability not to infect while time $dt$. So
probability to infect while time $dt$ is:
$$1 - (1-c)^{\frac{rI}{N}dt} = 1 - \exp(\frac{r\log(1-c)I}{N}dt)
\approx 1-(1+\frac{r\log(1-c)I}{N}dt) = -\frac{r\log(1-c)I}{N}dt$$
So $p(t) = p(I(t)) = -\frac{rI\log(1-c)}{N}$.

To create model of epidemic flow, it is need to solve differential equations
system with numeric method. To do so I used R package ```deSolve```, function 
```ode(y, times, func, parms, ...)```. Also, got values are decimals, which is
not correct to represent people, so I rounded $S, I$, floored $R, D$ and got
$J = N-(S+I+R+D)$.

As model parameters will be taken:
* $N$ - number of people
* $I_0$ - number of infected people at moment $t=0$, $0<I_0<N$
* $r$ - mean number of contacts per day for a single person
* $c$ - probability of infection for single contact with infected
* $\alpha$ - probability to recover while one day for a single latent infected 
* $\beta$ - probability to die while one day for a single latent infected
* $a$ - probability to recover while one day for a single tested infected
* $b$ - probability to die while one day for a single tested infected

