---
title: "Numerical Solution of a Variation Problem"
author: "Daniel Paliura"
date: "3/29/2021"
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
source('model.R', echo = FALSE)
```

## Introduction

This document was created in addition to 'Control Theory' subject third
laboratory work 'Numerical Solution of a Variation Problem'. This document on
GitHub is [here](https://github.com/dPaliura/Control_theory_labs/blob/main/lab3/report.pdf).
It contains performance of the task according to variant.

## The Task

1) Chosen example (variant 1 as 15-2*7) of variation problem to solve: 
    $\int_{0}^{1}((x')^2 + x^2)dt \to \min, x_0=x(0)=0, x_N=x(1)=1$.
2) To choose number of approximation points $N$.
3) To code function $J_N(x_1, \cdots ,x_{N-1}$.
4) To choose optimization method (non-gradient or gradient) for minimization
    $J_N(x_1, \cdots ,x_{N-1}$.
5) To apply chosen method to minimize $J_N(x_1, \cdots ,x_{N-1}$ and build
    plot of function value dependence on iteration number.
6) To visualize got solution as part-line plot by points
    $(t_i, x_i), i=0 \cdots N$.
7) To compare visually (by plots) got numeric solution and analytic one.
8) To research approximation accuracy dependently on number of approximation
    points $N$.
9) To prepare the report.